
## Matched pairs

Some data: 

\centering{
  \includegraphics[height=0.7\textheight]{Screenshot_2019-04-26_13-41-29}
}


## Matched pairs data
- Data are comparison of 2 drugs for effectiveness at reducing pain.

     - 12 subjects (cases) were arthritis sufferers
     - Response is #hours of pain relief from each drug.
      
- In reading example, each child tried only one reading method.
- But here, each subject tried out both drugs, giving us two
measurements.

    - Possible because, if you wait long enough, one drug has no influence
over effect of other.
    - Advantage: focused comparison of drugs. Compare one drug with
another on same person, removes a lot of variability due to differences between people. 
    - Matched pairs, requires different analysis.
      
- Design: randomly choose 6 of 12 subjects to get drug A first, other 6
get drug B first.

## Paired t test: reading the data
Values aligned in columns:  

```{r}
my_url <- 
  "http://www.utsc.utoronto.ca/~butler/c32/analgesic.txt"
pain <- read_table(my_url)
```

## The data

```{r}
pain
```

## Paired *t*-test 

\small

```{r}
with(pain, t.test(druga, drugb, paired = T))
```

\normalsize

- P-value is 0.053. 
- Not quite evidence of difference between drugs.

## t-testing the differences

- Likewise, you can calculate the differences yourself and
do a 1-sample t-test on them.
- First calculate a column of differences:

\footnotesize

```{r}
(pain %>% mutate(diff=druga-drugb) -> pain)
```

\normalsize

## t-test on the differences
- then throw them into t.test, testing that the mean is zero, with
same result as before:
```{r}
with(pain, t.test(diff, mu=0))
```

## Assessing normality 
- 1-sample and 2-sample t-tests assume (each) group normally
distributed.
- Matched pairs analyses assume (theoretically) that differences
normally distributed.
- How to assess normality? A normal quantile plot.

## The normal quantile plot (of differences)


```{r, fig.height=4}
ggplot(pain,aes(sample=diff))+stat_qq()+stat_qq_line()
```

- Points should follow the straight line. Bottom left one way off, so
normality questionable here: outlier.

## What to do instead?


- Matched pairs $t$-test based on one sample of differences
- the differences not normal (enough)
- so do *sign test* on differences, null median 0:


```{r}
sign_test(pain, diff, 0)
```


## Comments

- no evidence of any difference between drugs (P-value 0.1460)
- in $t$-test, the low outlier difference pulled mean difference downward and made it look more negative than it should have been
- therefore, there really isn't any difference between the drugs.

