
## The normal quantile plot

- see that normal distributions of data (or being normal enough) important
- only tools we have to assess this are histograms and maybe boxplots
- a better tool is **normal quantile plot**:
  - plot data against what you expect if data actually normal
  - look for points to follow a straight line, at least approx
- `ggplot` code: `aes` `sample`; geoms `stat_qq` and `stat_qq_line` 
  
## Kids learning to read

```{r, echo=FALSE, message=FALSE}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/drp.txt"
kids <- read_delim(my_url," ")
kids
```


```{r}
ggplot(kids, aes(x = group, y = score)) + geom_boxplot()
```
Each group looks normal, or at least symmetric.

## Get the groups separately

```{r}
kids %>% filter(group == "t") -> treatment
kids %>% filter(group == "c") -> control
```

to check

```{r}
treatment %>% count(group)
control %>% count(group)
```

## The treatment group

```{r, fig.height=4.5}
ggplot(treatment, aes(sample = score)) + 
  stat_qq() + stat_qq_line()
```
only problem here is lowest value a little too low (mild outlier).

## Control group

```{r, fig.height=4}
ggplot(control, aes(sample = score)) + 
  stat_qq() + stat_qq_line()
```

This time, highest value a little too high, but again, no real problem with normality.

## Facetting more than one sample

Use the whole data set and facet by groups

```{r, fig.height=4.5}
ggplot(kids, aes(sample = score)) + 
  stat_qq() + stat_qq_line() + facet_wrap(~group)
```


## Blue Jays attendances, skewed to right

```{r, echo=FALSE, message=FALSE}
jays <- read_csv("jays15-home.csv")
```


```{r}
ggplot(jays, aes(x = attendance)) + geom_histogram(bins = 6)
```

## On a normal quantile plot

```{r, fig.height=3.5}
ggplot(jays, aes(sample = attendance)) + 
  stat_qq() + stat_qq_line()
```

   
- Attendances at low end too bunched up: skewed to right. 
- Right-skewness can also show up as highest values being too high, or as a curved pattern in the points.

## More normal quantile plots
- How straight does a normal quantile plot have to be?
- There is randomness in real data, so even a normal quantile plot from
normal data won’t look perfectly straight.
- With a small sample, can look not very straight even from normal
data.
- Looking for systematic departure from a straight line; random wiggles
ought not to concern us.
- Look at some examples where we know the answer, so that we can
see what to expect.

## Normal data, large sample

```{r set-seed, echo=F}
set.seed(457299)
```


```{r, fig.height=4.5}
d <- tibble(x=rnorm(200))
d
ggplot(d, aes(x=x)) + geom_histogram(bins=10)
```

## The normal quantile plot

```{r, fig.height=4.5}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```

## Normal data, small sample

```{r, echo=F}
set.seed(457299)
```

- Not so convincingly normal, but not obviously skewed:


```{r normal-small, fig.height=4.5}
d <- tibble(x=rnorm(20))
ggplot(d, aes(x=x)) + geom_histogram(bins=5)
```


## The normal quantile plot

Good, apart from the highest and lowest points being slightly off. I’d call
this good:


```{r, fig.height=4.5}
ggplot(d, aes(sample=x)) + stat_qq() + stat_qq_line()
```

## Chi-squared data, *df* = 10

Somewhat skewed to right:

```{r, fig.height=4.5}
d <- tibble(x=rchisq(100, 10))
ggplot(d,aes(x=x)) + geom_histogram(bins=10)
```



## The normal quantile plot

Somewhat opening-up curve:

```{r, fig.height=4.5}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```


## Chi-squared data, df = 3

Definitely skewed to right:

```{r chisq-small-df, fig.height=4.5}
d <- tibble(x=rchisq(100, 3))
ggplot(d, aes(x=x)) + geom_histogram(bins=10)
```



## The normal quantile plot

Clear upward-opening curve:

```{r, fig.height=4.5}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```



## t-distributed data, df = 3

Long tails (or a very sharp peak):

```{r t-small, fig.height=4.5}
d <- tibble(x=rt(300, 3))
ggplot(d, aes(x=x)) + geom_histogram(bins=10)
```



## The normal quantile plot

Low values too low and high values too high for normal.

```{r, fig.height=4.5}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```

## Summary

On a normal quantile plot:

- points following line (with some small wiggles): normal.
- kind of deviation from a straight line indicates kind of nonnormality:
  - a few highest point(s) too high and/or lowest too low: outliers
  - else see how points at each end off the line:
  
  \vspace{2ex}
  
  \begin{center}
  
  \begin{tabular}{lcc}
  \hline
  & \multicolumn{2}{c}{High points}\\
  Low points & Too low & too high\\
  \hline
  Too low & Skewed left & Long tails\\
  Too high & Short tails & Skewed right \\
  \hline
  
  \end{tabular}
  
  \end{center}
  
  \vspace{2ex}
  
- short-tailed distribution OK for $t$ (mean still good), but others problematic (depending on sample size).  