
## Matched pairs

Some data: 

\centering{
  \includegraphics[height=0.7\textheight]{Screenshot_2019-04-26_13-41-29}
}


## Matched pairs data
- Data are comparison of 2 drugs for effectiveness at reducing pain.

     - 12 subjects (cases) were arthritis sufferers
     - Response is #hours of pain relief from each drug.
      
- In reading example, each child tried only one reading method.
- But here, each subject tried out both drugs, giving us two
measurements.

    - Possible because, if you wait long enough, one drug has no influence
over effect of other.
    - Advantage: focused comparison of drugs. Compare one drug with
another on same person, removes a lot of variability due to differences between people. 
    - Matched pairs, requires different analysis.
      
- Design: randomly choose 6 of 12 subjects to get drug A first, other 6
get drug B first.

## Paired t test: reading the data
Values aligned in columns:  

```{r}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/analgesic.txt"
pain <- read_table(my_url)
```

## The data

```{r}
pain
```

## Paired *t*-test 

\small

```{r}
with(pain, t.test(druga, drugb, paired = TRUE))
```

\normalsize

- P-value is 0.053. 
- Not quite evidence of difference between drugs.

## t-testing the differences

- Likewise, you can calculate the differences yourself and
do a 1-sample t-test on them.
- First calculate a column of differences:

\footnotesize

```{r}
(pain %>% mutate(diff=druga-drugb) -> pain)
```

\normalsize

## t-test on the differences
- then throw them into t.test, testing that the mean is zero, with
same result as before:
```{r}
with(pain, t.test(diff, mu=0))
```

## Assessing normality 
- 1-sample and 2-sample t-tests assume (each) group normally
distributed.
- Matched pairs analyses assume (theoretically) that differences
normally distributed.
- Though we know that t-tests generally behave well even without
normality.
- How to assess normality? A normal quantile plot.
  - Idea: scatter of points should follow the straight line, without curving.
  - Outliers show up at bottom left or top right of plot as points off the
line.

## The normal quantile plot

- of differences from matched pairs data

```{r, fig.height=3.3}
ggplot(pain,aes(sample=diff))+stat_qq()+stat_qq_line()
```

- Points should follow the straight line. Bottom left one way off, so
normality questionable here: outlier.

## More normal quantile plots
- How straight does a normal quantile plot have to be?
- There is randomness in real data, so even a normal quantile plot from
normal data won’t look perfectly straight.
- With a small sample, can look not very straight even from normal
data.
- Looking for systematic departure from a straight line; random wiggles
ought not to concern us.
- Look at some examples where we know the answer, so that we can
see what to expect.

## Normal data, large sample

```{r set-seed, echo=F}
set.seed(457299)
```


```{r, fig.height=3.5}
d=tibble(x=rnorm(200))
ggplot(d,aes(x=x))+geom_histogram(bins=10)
```

## The normal quantile plot

```{r, fig.height=3.8}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```

## Normal data, small sample

```{r, echo=F}
set.seed(457299)
```

- Not so convincingly normal, but not obviously skewed:


```{r normal-small, fig.height=3.1}
d=tibble(x=rnorm(20))
ggplot(d,aes(x=x))+geom_histogram(bins=5)
```


## The normal quantile plot

Good, apart from the highest and lowest points being slightly off. I’d call
this good:


```{r, fig.height=3.1}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```

## Chi-squared data, *df* = 10

Somewhat skewed to right:

```{r, fig.height=3.1}
d=tibble(x=rchisq(100,10))
ggplot(d,aes(x=x))+geom_histogram(bins=10)
```



## The normal quantile plot

Somewhat opening-up curve:

```{r, fig.height=3.3}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```


## Chi-squared data, df = 3

Definitely skewed to right:

```{r chisq-small-df, fig.height=3.1}
d=tibble(x=rchisq(100,3))
ggplot(d,aes(x=x))+geom_histogram(bins=10)
```



## The normal quantile plot

Clear upward-opening curve:

```{r, fig.height=3.4}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```



## t-distributed data, df = 3

Long tails (or a very sharp peak):

```{r t-small, fig.height=3.2}
d=tibble(x=rt(300,3))
ggplot(d,aes(x=x))+geom_histogram(bins=10)
```



## The normal quantile plot

Low values too low and high values too high for normal.

```{r, fig.height=3.4}
ggplot(d,aes(sample=x))+stat_qq()+stat_qq_line()
```


## Our pain-relief data

```{r pain-relief-qq, fig.height=3.8}
ggplot(pain,aes(sample=diff))+stat_qq()+stat_qq_line()
```

## Comments

- Definitely not normal. What to do?
- Sign test on differences, null median 0.

## Sign test
- Most easily: calculate differences in data frame, then use `smmr`.
- Null median difference is 0:

```{r}
library(smmr)
pain %>% mutate(mydiff=druga-drugb) %>%
  sign_test(mydiff,0)
```



## Comments 

- P-value 0.1460. No evidence that the drugs are different.
- Since we are working in a pipeline, input data frame to `sign_test` is
“whatever came out of previous step”.

## blue jays data

```{r}
my_url <- "http://ritsokiguess.site/STAC32/jays15-home.csv"
jays <- read_csv(my_url)
jays
```

## normal quantile plot of attendances

```{r}
ggplot(jays, aes(sample = attendance)) + stat_qq() + stat_qq_line()
```

## histogram

```{r}
ggplot(jays, aes(x = attendance)) + geom_histogram(bins=6)
```


## (Some of) the kids’ reading data, again

```{r, echo=FALSE}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/drp.txt"
kids <- read_delim(my_url," ")
```

```{r}
kids 
```

## Where we were at 

- 21 kids in “treatment”, new reading method; 23 in “control”,
standard reading method.

- Assessing assumptions:
  - We did two-sample t-test (Satterthwaite-Welch) before.
  - Assumes approx. normal data within each group.
  - Does not assume equal spread.
  - (Pooled t-test *does* assume equal spread).
  - Assess each group separately. 

## Boxplots for reading data

```{r, fig.height=3.7}
ggplot(kids,aes(x=group,y=score))+geom_boxplot()
```

## Facetted normal quantile plots
Done this way:

```{r, fig.height=3.2}
ggplot(kids,aes(sample=score))+stat_qq()+stat_qq_line()+
facet_wrap(~group)
```

## Comments
- These plots show no problems with normality. Both groups are more
or less symmetric/normal and there are no outliers.
- Equal spreads questionable, but we don’t need that.
- Assess equal spreads by looking at *slopes* of normal quantile plots.
- We ought be happy with the (Welch) two-sample t-test (over)


## Welch two-sample test

```{r}
t.test(score~group,data=kids,alternative="less")
```

from which we concluded that the new reading method really does
help.
