## Comparing models

- Typically in ANOVA situation, we want to see whether the group means are "really" different.
- We are used to doing an $F$-test here.
- In regression, have looked at AIC: measure of fit allowing for complexity.
- Bayesian equivalent called `looic`, in package `loo`.

## Difficulty:

- Posterior proportional to likelihood times prior:
  - Stan drops proportionality constant
  - but for model comparison, it matters.
- Have to calculate log-likelihood again, not dropping constants this time.

## The extra code

```
generated quantities {
  vector[n_obs] log_lik;
  int g;
  for (i in 1:n_obs) {
    g=group_no[i];
    log_lik[i] = normal_lpdf(density[i] | mu[g], sigma);
  }
}
```

## Comments

- This is repeat of `model` section, but note no "squiggle": this explicitly calculates the log of the normal density function for each observation and saves it in an array.
- This section goes at the bottom.

## Compile and sample

```{r, results="hide"}
anova_loo_compiled <- stan_model("anova-loo.stan")
anova_data <- list(
  n_obs = 30,
  n_group = 3,
  density = rats$density,
  group_no = rats$group_no
)
anova_loo_samples <- sampling(anova_loo_compiled, data = anova_data)
```

## Now we need a null model

- one value of `mu` for all groups
- replace all the `mu[i]` with `mu`
- omit any reference to group numbers (not needed any more)
- leave the `data` section as is (so can use previous `anova_data`).

## The null Stan code 1/2

```
data {
  int n_obs;
  int n_group;
  real density[n_obs];
  int<lower=1, upper=n_group> group_no[n_obs];
}

parameters {
  real mu;
  real<lower=0> sigma;
}

```

## The null Stan code 2/2

```
model {
  // priors
  mu ~ normal(600, 50);
  sigma ~ chi_square(50);
  // likelihood
  for (i in 1:n_obs) {
    density[i] ~ normal(mu, sigma);
  }
}

generated quantities {
  vector[n_obs] log_lik;
  for (i in 1:n_obs) {
    log_lik[i] = normal_lpdf(density[i] | mu, sigma);
  }
}
```

## Compile and sample again

```{r, results="hide"}
anova_loo_null_compiled <- stan_model("anova_loo_null.stan")
anova_loo_null_samples <- sampling(anova_loo_null_compiled, data = anova_data)
```

## Compare the fits of the two models


Setup

```{r}
library(loo)
log_lik_a <- extract_log_lik(anova_loo_samples,
  merge_chains = F
)
log_lik_0 <- extract_log_lik(anova_loo_null_samples,
  merge_chains = F
)
r_eff_a <- relative_eff(log_lik_a)
r_eff_0 <- relative_eff(log_lik_0)
```

## Results 1/2

\footnotesize
```{r}
loo(log_lik_a, r_eff = r_eff_a)
```
\normalsize

## Results 2/2

\small
```{r}
loo(log_lik_0, r_eff = r_eff_0)
```
\normalsize


## Comments

- Look at `looic`: smaller value is better, allowing for complexity of model
- For separate means, 277.7; for one single mean, 285.7.
- Prefer the model with separate means, one for each group.

