## Packages for this section

```{r, eval=F}
library(tidyverse)
library(rstan)
```


## Bayesian and frequentist inference

- The inference philosophy that we have learned so far says that:
  - parameters to be estimated are *fixed* but *unknown*
  - Data random; if we took another sample we'd get different data.
- This is called "frequentist" or "repeated-sampling" inference.
- Bayesian inference says:
  - *parameters* are random, *data* is *given*
- Ingredients:
  - **prior distribution**: distribution of parameters before seeing data.
  - **likelihood**: model for data if the parameters are known 
  - **posterior distribution**: distribution of parameters *after* seeing data.
  
## Distribution of parameters

- Instead of having a point or interval estimate of a parameter, we have an entire distribution
- so in Bayesian statistics we can talk about eg.
  - probability that a parameter is bigger than some value
  - probability that a parameter is close to some value
  - probability that one parameter is bigger than another
  
- Name comes from Bayes' Theorem, which here says

> posterior is proportional to likelihood times prior

- more discussion about this is in 
[**a blog post**](http://ritsokiguess.site/docs/2018/02/28/working-my-way-back-to-you-a-re-investigation-of-rstan/). 

## An example

- Suppose we have these (integer) observations:

```{r}
(x <- c(0, 4, 3, 6, 3, 3, 2, 4))
```

- Suppose we believe that these come from a Poisson distribution with a mean $\lambda$ that we want to estimate.
- We need a prior distribution for $\lambda$. I will (for some reason) take a $Weibull$ distribution with parameters 1.1 and 6, that has quartiles 2 and 6. Normally this would come from your knowledge of the data-generating *process*.
- The Poisson likelihood can be written down (see over).

## Some algebra

- We have $n=8$ observations $x_i$, so the Poisson likelihood is proportional to

$$ \prod_{i=1}^n e^{-\lambda} \lambda^{x_i} = e^{-n\lambda} \lambda^S, $$
where $S=\sum_{i=1}^n x_i$. 

- then you write the Weibull prior density (as a function of $\lambda$):

$$ C (\lambda/6)^{0.1} e^{-(\lambda/6)^{1.1}}  $$
where $C$ is a constant.

- and then you multiply these together and try to recognize the distributional form. Only, here you can't. The powers 0.1 and 1.1 get in the way.

## Sampling from the posterior distribution

- Wouldn't it be nice if we could just *sample* from the posterior distribution? Then we would be able to compute it as accurately as we want.

- Metropolis and Hastings: devise a Markov chain (C62) whose limiting distribution is the posterior you want, and then sample from that Markov chain (easy), allowing enough time to get close enough to the limiting distribution.

- Stan: uses a modern variant that is more efficient (called Hamiltonian Monte Carlo), implemented in R package `rstan`. 

- Write Stan code in a file, compile it and sample from it.

## Components of Stan code: the model

```
model {
// likelihood
x ~ poisson(lambda);
}
```

This is how you say "$X$ has a Poisson distribution with mean $\lambda$". **Note that lines of Stan code have semicolons on the end.**

## Components of Stan code: the prior distribution

```
model {
// prior
lambda ~ weibull(1.1, 6);
// likelihood
x ~ poisson(lambda);
}
```

## Components of Stan code: data and parameters (first in the Stan code)

```
data {
int x[8];
}

parameters {
real<lower=0> lambda;
}
```

## Compile and sample from the model

```{r, eval=F}
poisson1_code <- stan_model(file = "poisson1.stan")
```

- set up data

```{r}
poisson1_data <- list(x = x)
```

- sample

```{r, eval=F}
poisson1_fit <- sampling(poisson1_code, data = poisson1_data)
```

```{r, include=F}
poisson1_code <- readRDS("poisson1_code.rds")
```


```{r, echo=F, results="hide"}
set.seed(457299)
poisson1_data <- list(x = x)
poisson1_fit <- sampling(poisson1_code, data = poisson1_data)
```

```{r, include=F}
saveRDS(poisson1_code, "poisson1_code.rds")
```

## The output

```{r}
poisson1_fit
```


## Comments

- This summarizes the posterior distribution of $\lambda$
- the posterior mean is 3.20 
- with a 95% posterior interval of 2.10 to 4.56. 
- The probability that $\lambda$ is between these two values really is 95%.

## Making the code more general

- The coder in you is probably offended by hard-coding the sample size and the parameters of the prior distribution. More generally:

```
data {
  int<lower=1> n;
  real<lower=0> a;
  real<lower=0> b;
  int x[n];
}
...
model {
// prior
lambda ~ weibull(a, b);
// likelihood
x ~ poisson(lambda);
}
```

## Set up again and sample:

- Compile again:

```{r, include=F}
poisson2_code <- readRDS("poisson2_code.rds")
poisson2_code
```


```{r, eval=F}
poisson2_code <- stan_model(file = "poisson2.stan")
```

```{r, include=F}
saveRDS(poisson2_code, "poisson2_code.rds")
```


- set up the data again including the new things we need:

```{r}
poisson2_data <- list(x = x, n = length(x), a = 1.1, b = 6)
```

- sample again

```{r, results="hide"}
poisson2_fit <- sampling(poisson1_code, data = poisson2_data)
```


## output should be the same (to within randomness)

```{r}
poisson2_fit
```

## Extracting actual sampled values

- `rstan` has `extract` for this. There is also an `extract` in `dplyr`: make sure you have the right one.

```{r}
poisson2_out <- extract(poisson2_fit)
poisson2_out
ggplot(tibble(lambda = poisson2_out$lambda), aes(x = lambda)) +
  geom_histogram(bins = 20)
ggplot(tibble(lambda = poisson2_out$lambda), aes(sample=lambda)) + stat_qq() + stat_qq_line()
```

## Posterior predictive distribution

- Another use for the actual sampled values is to see what kind of *response* values we might get in the future. This should look something like our data. For a Poisson distribution, the response values are integers:

```{r, fig.height=3}
tibble(lambda = poisson2_out$lambda) %>%
  mutate(x_sim = map_int(lambda, ~ rpois(1, .))) -> d
```

## A bar chart:

```{r}
ggplot(d, aes(x = x_sim)) + geom_bar()
```


## Comparison

Our actual data values were these:

```{r}
x
```

- None of these are very unlikely according to our posterior predictive distribution, so our model is believable. 
- Or make a plot: a bar chart with the data on it as well (over):

```{r}
ggplot(d, aes(x = x_sim)) + geom_bar() +
  geom_dotplot(data = tibble(x), aes(x = x), binwidth = 1) +
  scale_y_continuous(NULL, breaks = NULL) -> g
```

- This also shows that the distribution of the data conforms well enough to the posterior predictive distribution (over).

## The plot 

```{r}
g
```

## Do they have the same distribution?

```{r}
qqplot(d$x_sim, x, plot.it = F) %>% as_tibble() -> dd
ggplot(dd, aes(x=x, y=y)) + geom_point()
```


the observed zero is a bit too small compared to expected (from the posterior), but the other points seem pretty well on a line.

## Analysis of variance, the Bayesian way

Recall the jumping rats data:

```{r, message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/jumping.txt"
rats0 <- read_delim(my_url, " ")
rats0
```

## Our aims here

- Estimate the mean bone density of all rats under each of the experimental conditions
- Model: given the group means, each observation normally distributed with common variance $\sigma^2$
- Three parameters to estimate, plus the common variance.
- Obtain posterior distributions for the group means.
- Ask whether the posterior distributions of these means are sufficiently different.

## Numbering the groups

- Stan doesn't handle categorical variables (everything is `real` or `int`).
- Turn the groups into group *numbers* first.
- Take opportunity to put groups in logical order:

```{r}
rats0 %>% mutate(
  group_fct = fct_inorder(group),
  group_no = as.integer(group_fct)
) -> rats
rats
```

## Plotting the data 1/2

Most obviously, boxplots:

```{r, fig.height=3.4}
ggplot(rats, aes(x = group_fct, y = density)) + geom_boxplot()
```

## Plotting the data 2/2

Another way: density plot (smoothed out histogram); can distinguish groups by colours:

```{r density_plot, fig.height=3.1}
ggplot(rats, aes(x = density, colour = group_fct)) +
  geom_density()
```


## The procedure

- For each observation, find out which (numeric) group it belongs to, 
- then model it as having a normal distribution with that group's mean and the common variance.
- Stan does `for` loops.

## The model part

Suppose we have `n_obs` observations:

```
model {
  // likelihood
  for (i in 1:n_obs) {
    g=group_no[i];
    density[i] ~ normal(mu[g], sigma);
  }
}
```

## The variables here

- `n_obs` is data.
- `g` is a temporary integer variable only used here
- `i` is only used in the loop (integer) and does not need to be declared
- `density` is data, a real vector of length `n_obs`
- `mu` is a parameter, a real vector of length 3 (3 groups)
- `sigma` is a real parameter

`mu` and `sigma` need prior distributions: 

  - for `mu`, each component independently normal with mean 600 and SD 50 (my guess at how big and variable they will be)
  - for `sigma`, chi-squared with 50 df (my guess at typical amount of variability from obs to obs)

## Complete the `model` section:

```
model {
  int g;
  // priors
  mu ~ normal(600, 50);
  sigma ~ chi_square(50);
  // likelihood
  for (i in 1:n_obs) {
    g=group_no[i];
    density[i] ~ normal(mu[g], sigma);
  }
}
```

## Parameters

The elements of `mu`, one per group, and also `sigma`, scalar:

```
parameters {
  real mu[n_group];
  real<lower=0> sigma;
}
```

- `sigma` has to be positive. Declare it so here, so that the sampling runs smoothly.
- declare `n_group` in data section 

## Data

Everything else:

```
data {
  int n_obs;
  int n_group;
  real density[n_obs];
  int<lower=1, upper=n_group> group_no[n_obs];
}
```

## Compile

Arrange these in order data, parameters, model in file `anova.stan`, then:

```{r, eval=F}
anova_compiled <- stan_model("anova.stan")
```

```{r read_anova_compiled, echo=FALSE}
anova_compiled <- readRDS("anova_compiled.rds")
```

## Set up data and sample

Supply values for *everything* declared in `data`: 

```{r anova_data_samples}
anova_data <- list(
  n_obs = 30,
  n_group = 3,
  density = rats$density,
  group_no = rats$group_no
)
anova_samples <- sampling(anova_compiled, data = anova_data)
```

## Check that the sampling worked properly 

```{r, fig.height=4}
traceplot(anova_samples)
```

## Comments

- The sampled values for each of the parameters should move freely across their posterior distributions (and not get stuck anywhere). 
- This appears to have happened.

## Look at the results

```{r anova_samples}
anova_samples
```

## Comments

- The posterior 95% intervals for control (group 1) and highjump (group 3) do not quite overlap, suggesting that these exercise groups really are different.
- Bayesian approach does not normally do tests: look at posterior distributions and decide whether they are different enough to be worth treating as different.

## Plotting the posterior distributions for the `mu`

- Extract the sampled `mu` values (matrix):

```{r anova_ext}
anova_ext <- extract(anova_samples)
anova_ext$mu %>% as_tibble()
```

```{r}
anova_ext$mu %>% as_tibble() %>% 
  count(V3>V1)
```


## Turn into a data frame, arrange for plotting, name groups

```{r, warning=F}
cbind(anova_ext$mu, sigma = anova_ext$sigma) %>%
  as_tibble() %>%
<<<<<<< HEAD
  pivot_longer(V1:V3, names_to="group", values_to="density") %>%
=======
  pivot_longer(starts_with("V"), names_to = "group", values_to = "density") %>%
>>>>>>> 2991ae36dff40038e717965e9023aeadbc00f334
  mutate(group = fct_recode(
    group,
    Control = "V1",
    Lowjump = "V2",
    Highjump = "V3"
  )) -> sims
sims
```

## What we have now:

```{r}
sims %>% sample_n(8)
```


## Density plots of posterior mean distributions

```{r, fig.height=4}
ggplot(sims, aes(x = density, colour = group)) + geom_density()
```

## Posterior predictive distributions

Randomly sample from posterior means and SDs in `sims`. There are 12000 rows in `sims`:

```{r}
sims %>% mutate(sim_data = rnorm(12000, density, sigma)) -> ppd
ppd
```

## Compare posterior predictive distribution with actual data

- Check that the model works: distributions of data similar to what we'd predict
- Idea: make plots of posterior predictive distribution, and plot actual data as points on them
- Use facets, one for each treatment group: 

```{r ppdgraph}
my_binwidth <- 15
ggplot(ppd, aes(x = sim_data)) +
  geom_histogram(binwidth = my_binwidth) +
  geom_dotplot(
    data = rats, aes(x = density),
    binwidth = my_binwidth
  ) +
  facet_wrap(~group) +
  scale_y_continuous(NULL, breaks = NULL) -> g
```

- See (over) that the data values are mainly in the middle of the predictive distributions.
- Even the control group that had outliers. 

## The plot

```{r, fig.height=3.75}
g
```

## Extensions 

- if you want a different model other than normal, change distribution in `model` section
- if you want to allow unequal spreads, create `sigma[n_group]` and in model `density[i] ~ normal(mu[g], sigma[g]);`
- Stan will work just fine after you recompile
- very flexible.
- Typical modelling strategy: start simple, add complexity as warranted by data.

## Comparing models

- Typically in ANOVA situation, we want to see whether the group means are "really" different.
- We are used to doing an $F$-test here.
- In regression, have looked at AIC: measure of fit allowing for complexity.
- Bayesian equivalent called `looic`, in package `loo`.

## Difficulty:

- Posterior proportional to likelihood times prior:
  - Stan drops proportionality constant
  - but for model comparison, it matters.
- Have to calculate log-likelihood again, not dropping constants this time.

## The extra code

```
generated quantities {
  vector[n_obs] log_lik;
  int g;
  for (i in 1:n_obs) {
    g=group_no[i];
    log_lik[i] = normal_lpdf(density[i] | mu[g], sigma);
  }
}
```

## Comments

- This is repeat of `model` section, but note no "squiggle": this explicitly calculates the log of the normal density function for each observation and saves it in an array.
- This section goes at the bottom.

## Compile and sample

```{r, results="hide"}
anova_loo_compiled <- stan_model("anova-loo.stan")
anova_data <- list(
  n_obs = 30,
  n_group = 3,
  density = rats$density,
  group_no = rats$group_no
)
anova_loo_samples <- sampling(anova_loo_compiled, data = anova_data)
```

## Now we need a null model

- one value of `mu` for all groups
- replace all the `mu[i]` with `mu`
- omit any reference to group numbers (not needed any more)
- leave the `data` section as is (so can use previous `anova_data`).

## The null Stan code 1/2

```
data {
  int n_obs;
  int n_group;
  real density[n_obs];
  int<lower=1, upper=n_group> group_no[n_obs];
}

parameters {
  real mu;
  real<lower=0> sigma;
}

```

## The null Stan code 2/2

```
model {
  // priors
  mu ~ normal(600, 50);
  sigma ~ chi_square(50);
  // likelihood
  for (i in 1:n_obs) {
    density[i] ~ normal(mu, sigma);
  }
}

generated quantities {
  vector[n_obs] log_lik;
  for (i in 1:n_obs) {
    log_lik[i] = normal_lpdf(density[i] | mu, sigma);
  }
}
```

## Compile and sample again

```{r, results="hide"}
anova_loo_null_compiled <- stan_model("anova_loo_null.stan")
anova_loo_null_samples <- sampling(anova_loo_null_compiled, data = anova_data)
```

## Compare the fits of the two models


Setup

```{r}
library(loo)
log_lik_a <- extract_log_lik(anova_loo_samples,
  merge_chains = F
)
log_lik_0 <- extract_log_lik(anova_loo_null_samples,
  merge_chains = F
)
r_eff_a <- relative_eff(log_lik_a)
r_eff_0 <- relative_eff(log_lik_0)
```

## Results 1/2

\footnotesize
```{r}
loo(log_lik_a, r_eff = r_eff_a)
```
\normalsize

## Results 2/2

\small
```{r}
loo(log_lik_0, r_eff = r_eff_0)
```
\normalsize


## Comments

- Look at `looic`: smaller value is better, allowing for complexity of model
- For separate means, 277.7; for one single mean, 285.7.
- Prefer the model with separate means, one for each group.

