## Tidying data
- Data rarely come to us as we want to use them.
- Before we can do analysis, typically have organizing to do.
- This is typical of ANOVA-type data, “wide format”:

```
     pig feed1 feed2 feed3 feed4
       1  60.8  68.7  92.6  87.9
       2  57.0  67.7  92.1  84.2
       3  65.0  74.0  90.2  83.1
       4  58.6  66.3  96.5  85.7
       5  61.7  69.8  99.1  90.3
```

- 20 pigs randomly allocated to one of four feeds. At end of 
study, weight of each pig is recorded. 
- Are any differences in mean weights among the feeds?
- Problem: want all weights in one column, with 2nd column
labelling which feed. Untidy!


## Tidy and untidy data (Wickham) 
- Data set easier to deal with if:
  - each observation is one row
  - each variable is one column
  - each type of observation unit is one table
- Data arranged this way called “tidy”; otherwise called “untidy”.
- For the pig data:
  - response variable is weight, but scattered over 4
columns, which are levels of a factor `feed`.
  - Want all the weights in one column, with a second column `feed`
saying which feed that weight goes with.
  - Then we can run `aov`.
  
## Packages for this section

```{r, eval=F}
library(tidyverse)
library(readxl)
```

  
## Reading in the pig data

```{r, message=F}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/pigs1.txt"
pigs1 <- read_delim(my_url, " ")
pigs1
```

## Gathering up the columns (OLD)
- This is a very common reorganization, and the magic “verb” is
`pivot_longer`:
```{r}
pigs1 %>% pivot_longer(feed1:feed4, names_to="feed", 
                       values_to="weight") -> pigs2
```

- `pigs2` is now in “long” format, ready for analysis. See next page.
- Anatomy of `pivot_longer`: 
    - columns to combine
    - a name for column that will contain groups
    - a name for column that will contain measurements

## Long format pigs

```{r}
pigs2
```

<<<<<<< HEAD
## Another way to do this: (NEW)

```{r}
pigs1 %>% 
  pivot_longer(feed1:feed4, names_to = "feed", 
               values_to="weight") -> pigs2
pigs2
=======
## Identifying the pigs

- Values in `pig` identify pigs *within each group*: pig 1 is four different pigs!
- Create unique pig IDs by gluing pig number onto feed:

```{r}
pigs2 %>% mutate(pig_id=str_c(feed, "_", pig)) -> pigs2
pigs2 %>% slice_sample(n=7)
>>>>>>> 2991ae36dff40038e717965e9023aeadbc00f334
```


## ...and finally, the analysis
- which is just what we saw before:
```{r}
weight.1 <- aov(weight ~ feed, data = pigs2)
summary(weight.1)
```

- The mean weights of pigs on the different feeds are definitely not all
equal.
- So we run Tukey to see which ones differ (over).

## Tukey

```{r}
TukeyHSD(weight.1)
```

All of the feeds differ! 

## Mean weights by feed 

To find the best and worst, get mean weight by 
feed group. I borrowed an idea from later to put the means in descending order:

```{r}
pigs2 %>%
  group_by(feed) %>%
  summarize(mean_weight = mean(weight))%>%
  arrange(desc(mean_weight))
```

Feed 3 is best, feed 1 worst.

## Should we have any concerns about the ANOVA?


```{r pigfeedplot}
ggplot(pigs2, aes(x = feed, y = weight)) + geom_boxplot()
```

## Comments

- Feed 2 has an outlier
- But there are only 5 pigs in each group
- The
conclusion is so clear that I am OK with this.


## Tuberculosis

- The World Health Organization keeps track of number of cases of
various diseases, eg. tuberculosis.
- Some data:

```{r}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/tb.csv"
tb <- read_csv(my_url)
```


## The data

<<<<<<< HEAD
```{r}
tb
=======
```{r glimpsetb}
glimpse(tb)
>>>>>>> 2991ae36dff40038e717965e9023aeadbc00f334
```

## What we have

- Variables:  country (abbreviated), year. Then number of
cases for each gender and age group, eg. `m1524` is males aged 15–24.
Also `mu` and `fu`, where age is unknown.
- Lots of missings. Want to get rid of.
- Abbreviations [here](https://en.wikipedia.org/wiki/ISO_3166-1_alpha-2).


<<<<<<< HEAD
OLD
```{r}
tb %>% gather(genage, freq, m04:fu, na.rm = T) -> tb2
```
=======
>>>>>>> 2991ae36dff40038e717965e9023aeadbc00f334


## All frequencies in one column

NEW

```{r}
tb %>% 
  pivot_longer(m04:fu, names_to = "genage", 
               values_to = "freq", values_drop_na = T) -> tb2
```

- columns to make longer
- column to contain the names
- column to contain the values
- (optional) drop missings in the values

## Results (some)

```{r}
tb2
```

## Separating
- 4 columns, but 5 variables, since `genage` contains both gender and
age group. Split that up using separate.
- `separate` needs 3 things:
  - what to separate (no quotes needed),
  - what to separate into (here you do need quotes),
  - how to split.
- For “how to split”, here “after first character”:
```{r}
tb2 %>% separate(genage, c("gender", "age"), 1) -> tb3
```

## Tidied tuberculosis data (some)

```{r}
tb3
```

```{r, echo=FALSE}
write_rds(tb3, "tb3.rds")
```


## In practice...

- instead of doing the pipe one step at a time, you *debug* it one step at
a time, and when you have each step working, you use that step’s
output as input to the next step, thus:
```{r}
tb %>%
  pivot_longer(m04:fu, names_to = "genage", 
               values_to = "freq", values_drop_na = T) %>% 
  separate(genage, c("gender", "age"), 1) 
```

- You can split the R code over as many lines as you like, as long as
each line is incomplete, so that R knows more is to come.
- I like to put the pipe symbol on the end of the line.

## Total tuberculosis cases by year (some of the years)

```{r}
tb3 %>%
  filter(between(year, 1991, 1998)) %>% 
  group_by(year) %>% summarize(total=sum(freq))
```

- Something very interesting happened between 1994 and 1995.

## To find out what

- try counting up total cases by country:

```{r}
tb3 %>% group_by(iso2) %>% 
  summarize(total=sum(freq)) %>% 
  arrange(desc(total))
```

## what years do I have for China?

China started recording in 1995, which is at least part of the problem:

```{r}
tb3 %>% filter(iso2=="CN") %>% 
  group_by(year) %>% 
  summarize(total=sum(freq))
```

## first year of recording for each country?

- A lot of countries started recording in about 1995, in fact:

```{r, echo=FALSE}
options(dplyr.summarise.inform = FALSE)
```

\small

```{r}
tb3 %>% group_by(iso2) %>% 
  summarize(first_year=min(year)) %>% 
  count(first_year)
```

\normalsize

## Some Toronto weather data

```{r}
my_url <- 
  "http://ritsokiguess.site/STAC32/toronto_weather.csv"
weather <- read_csv(my_url)
```

## The data (some)

```{r}
weather
```

## The columns

- Daily weather records for "Toronto City" weather station in 2018:

  - *station*: identifier for this weather station (always same here)
  - *Year*, *Month*
  - *element*: whether temperature given was daily max or daily min
  - *d01, d02*,... *d31*: day of the month from 1st to 31st.

- Numbers in data frame all temperatures (for different days of the month),
so first step is

```{r}
weather %>% 
  pivot_longer(d01:d31, names_to="day", 
               values_to="temperature", 
               values_drop_na = T) -> d
```

## So far

```{r}
d
```

## The days
- Column `element` contains names of two different variables, that
should each be in separate column.
- Distinct from eg. `m1524` in tuberculosis data, that contained levels of
two different factors, handled by separate.
- Untangling names of variables handled by `pivot_wider`:

```{r}
weather %>%
  pivot_longer(d01:d31, names_to="day", 
               values_to="temperature", 
               values_drop_na = T) %>% 
  pivot_wider(names_from=element, 
                values_from=temperature) -> d
```

## So far

```{r}
d
```

## Further improvements
- We have tidy data now, but can improve things further.
- `mutate` creates new columns from old (or assign back to change a
variable).
- Would like numerical dates. `separate` works, or pull out number as
below.
- `select` keeps columns (or drops, with minus). Station name has no
value to us:

\small
```{r}
weather %>%
  pivot_longer(d01:d31, names_to="day", 
               values_to="temperature", values_drop_na = T) %>% 
  pivot_wider(names_from=element, values_from=temperature) %>% 
  mutate(Day = parse_number(day)) %>%
  select(-station) -> d
```
\normalsize
## So far

```{r}
d
```

## Final step(s)
- Make year-month-day into proper date.
- Keep only date, tmax, tmin:
\small
```{r}
weather %>%
  pivot_longer(d01:d31, names_to="day", 
               values_to="temperature", values_drop_na = T) %>% 
  pivot_wider(names_from=element, values_from=temperature) %>% 
  mutate(Day = parse_number(day)) %>%
  select(-station) %>% 
  unite(datestr, c(Year, Month, Day), sep = "-") %>%
  mutate(date = as.Date(datestr)) %>%
  select(c(date, tmax, tmin)) -> weather_tidy
```
\normalsize

## Our tidy data frame

```{r}
weather_tidy
```

## Plotting the temperatures
- Plot temperature against date joined by lines, but with separate lines
for max and min. `ggplot` requires something like

\footnotesize
```{r, eval=F}
ggplot(..., aes(x = date, y = temperature)) + geom_point() + 
  geom_line()
```
\normalsize

only we have two temperatures, one a max and one a min, that we
want to keep separate.

- The trick: combine `tmax` and `tmin` together into one column, keeping
track of what kind of temp they are. (This actually same format as
untidy `weather`.) Are making `weather_tidy`
untidy for purposes of drawing graph only.
- Then can do something like

\footnotesize
```{r, eval=F}
ggplot(d, aes(x = date, y = temperature, colour = maxmin)) 
  + geom_point() + geom_line()
```
\normalsize
to distinguish max and min on graph.

## Setting up plot
- Since we only need data frame for plot, we can do the
column-creation and plot in a pipeline.
- For a `ggplot` in a pipeline, the initial data frame is omitted, because it
is whatever came out of the previous step.
- To make those “one column”s: `pivot_longer`. I save the graph to show overleaf:
```{r}
weather_tidy %>%
  pivot_longer(tmax:tmin, names_to="maxmin", 
               values_to="temperature") %>%
  ggplot(aes(x = date, y = temperature, colour = maxmin)) +
      geom_line() -> g
```

## The plot

```{r temp-plot}
g
```

## Summary of tidying “verbs”

  \begin{tabular}{lp{0.65\textwidth}}
    Verb & Purpose\\
    \hline
    \texttt{pivot\_longer} & Combine columns that measure same thing into one\\
    \texttt{pivot\_wider} & Take column that measures one thing under
                     different conditions and put into multiple columns\\
    \texttt{separate} & Turn a column that encodes
                        several variables into
                        several columns\\
    \texttt{unite} & Combine several (related) variables into one
                     ``combination'' variable\\
    \hline
  \end{tabular}
  
  `pivot_longer` and `pivot_wider` are opposites; `separate` and
  `unite` are opposites.
  


