## Packages for this section

- This is (almost) all base R! We only need this for one thing later:

```{r, eval=F}
library(tidyverse)
```


## Vector addition


Adds 2 to each element.

- Adding vectors:
```{r}
u <- c(2, 3, 6, 5, 7)
u
v <- c(1, 8, 3, 2, 0)
u
v
u + v
```

- Elementwise addition. (Linear algebra: vector addition.)

## Adding a number to a vector

- Define a vector, then “add 2” to it:

```{r}
u
k <- 2
u
u + k
```

- adds 2 to *each* element of `u`.

this is as in a dataframe:

```{r}
d <- tribble(
  ~x, ~y,
  1, 2,
  3, 4
)
d %>% mutate(z = 5)
```


## Scalar multiplication
As per linear algebra:

```{r}
k
u
k * u
```

- Each element of vector multiplied by 2.

## “Vector multiplication”
What about this?
```{r}
u
v
u * v
```

Each element of `u` multiplied by *corresponding* element of `v`. Could be
called elementwise multiplication. 

(Don't confuse with “outer” or
“vector” product from linear algebra, or indeed “inner” or “scalar” multiplication,
for which the answer is a number.)

## Combining different-length vectors
- No error here (you get a warning). What happens?
```{r}
u
w <- c(1, 2)
w
u + w
```

- Add 1 to first element of `u`, add 2 to second.
- Go back to beginning of `w` to find something to add: add 1 to 3rd
element of `u`, 2 to 4th element, 1 to 5th. 

## How R does this 

- Keep re-using shorter vector until reach length of longer one.
- “Recycling”.
- If the longer vector's length not a multiple of the shorter vector's length, get a warning (probably not what you want). 
- Same idea is used when multiplying a vector by a number: the number
keeps getting recycled.

## Matrices
- Create matrix like this:
```{r}
(A <- matrix(1:4, nrow = 2))
```

- First: stuff to make matrix from, then how many rows and columns.
- R goes down columns by default. To go along rows instead:
```{r}
(B <- matrix(5:8, nrow = 2, ncol = 2, byrow = TRUE))
```

- One of `nrow` and `ncol` enough, since R knows how many things in
the matrix.

## Adding matrices
What happens if you add two matrices?

```{r}
A
B
A + B
```

## Adding matrices

- Nothing surprising here. This is matrix addition as we and linear algebra know it.

## Multiplying matrices
- Now, what happens here?
```{r}
A
B
A * B
```

## Multiplying matrices?

- *Not* matrix multiplication (as per linear algebra).
- Elementwise multiplication. Also called *Hadamard product* of `A` and `B`.

## Legit matrix multiplication
Like this:

```{r}
A
B
A %*% B
```

## Reading matrix from file
- The usual:
```{r}
my_url <- "http://www.utsc.utoronto.ca/~butler/c32/m.txt"
M <- read_delim(my_url, " ", col_names = F)
M
class(M)
```

## but...

- except that M is not an R matrix, and thus this doesn’t work:
```{r, error=T}
v <- c(1, 3)
v
M %*% v
```

## Making a genuine matrix

Do this first:
```{r}
M <- as.matrix(M)
M
```

and then all is good:
```{r}
M %*% v
```

## Linear algebra stuff
- To solve system of equations
$Ax = w$ for $x$:
```{r}
A
w
solve(A, w)
```

## Matrix inverse

- To find the inverse of A:
```{r}
A
solve(A)
A %*% solve(A)
```

- You can check that the matrix inverse and equation solution are
correct.

## Inner product
- Vectors in R are column vectors, so just do the matrix multiplication (`t()` is transpose): 

```{r}
a <- c(1, 2, 3)
b <- c(4, 5, 6)
t(a) %*% b
as.vector(t(a) %*% b)
```

- Note that the answer is actually a 1 × 1 matrix.
- Or as the sum of the elementwise multiplication:

```{r}
a
b
a*b
sum(a * b)
```

## Accessing parts of vector

- use square brackets and a number to get elements of a vector

```{r}
b
b[3]
```

## Accessing parts of matrix

- use a row and column index to get an element of a matrix

```{r}
A
A[2,1]
```

- leave the row or column index empty to get whole row or column, eg.

```{r}
A[1,]
A[,2]
```



## Eigenvalues and eigenvectors

- For a matrix $A$, these are scalars $\lambda$ and vectors $v$ that solve

$$ A v = \lambda v $$

- In R, `eigen` gets these:

```{r}
A
e <- eigen(A)
```

## The eigenvalues/vectors

```{r}
e
```


## To check that the eigenvalues/vectors are correct

- $\lambda_1 v_1$: (scalar) multiply first eigenvalue by first eigenvector (in column)

```{r}
e$values[1] * e$vectors[,1]
```
- $A v_1$: (matrix) multiply matrix by first eigenvector (in column)

```{r}
A %*% e$vectors[,1]
```

- These are (correctly) equal.
- The second one goes the same way.

## A statistical application of eigenvalues

- A negative correlation:

\footnotesize
```{r}
d <- tribble(
  ~x,  ~y,
  10,  20,
  11,  18,
  12,  17,
  13,  14,
  14,  13
)
d
v <- cor(d)
v
```
\normalsize

- `cor` gives the correlation matrix between each pair of variables (correlation between `x` and `y` is $-0.988$)

## Eigenanalysis of correlation matrix

```{r}
eigen(v)
```

- first eigenvalue much bigger than second (second one near zero)
- two variables, but data nearly *one*-dimensional
- opposite signs in first eigenvector indicate that the one dimension is:
  - `x` small and `y` large at one end,
  - `x` large and `y` small at the other.
